---
title: "ChipSortPaper_Fig3"
author: "Anthony R Cillo"
date: "2026-05-22"
output: github_document
---

## Load packages 

```{r}

library(readxl)
library(ggplot2)
library(ggpubr)
library(patchwork)
library(dplyr)
library(tidyverse)
library(qs)
library(Seurat)
library(here)

```

## Load data 

```{r}

df <- read_excel("../01_input/ChipSort_data_20260310.xlsx", sheet = "sort")

ser <- qs::qread("../01_input/ser_hd9.qs")

```

## Establish color scheme

```{r}

## Refined based on cell types refinements below
my_colors <- c(
"Classical Monocytes" = "#79C6C8",
"Inflamatory Monocytes" = "#F0A020",
"Intermediate Monocytes"= "#5E548E", 
"B cells" = "#E4571B", 
"NK cells" = "#5F9C94", 
"T cells" = "#9A8BB0", 
"Nonclassical Monocytes" = "#8FAF8F",
"Dendritic cells" = "#C97C6A",
"pDCs" = "#0F4C5C")


#color palette
colors_pal <- c(
"#79C6C8",
"#F0A020",
"#5E548E",
"#8FAF8F",
"#E4571B",
"#5F9C94",
"#9A8BB0",
"#C97C6A",
"#0F4C5C",
"#C9D8A6",
"#F06A4A",
"#6A2E8C",
"#5A7D4F",
"#D8A18B",
"#B7492A")


```

## Plot expected vs actual recovery from chip sorting 

```{r}

df_hd9 <- df 

df_long <- df_hd9 %>%
  pivot_longer(
    cols = c(target_recovery, pre_filt_recovery),
    names_to = "recovery_type",
    values_to = "recovery")

df_long$cells_loaded <- recode(df_long$cells_loaded,
                               `2500`="2,500",
                               `5000`="5,000",
                               `10000`="10,000",
                               `20000`="20,000")

df_long$cells_loaded <- factor(
  df_long$cells_loaded,
  levels=c("2,500","5,000","10,000","20,000")
)

df_long$recovery_type <- factor(
  df_long$recovery_type,
  levels = c("target_recovery", "pre_filt_recovery"))

p1 <- df_long %>%
  filter(sort_type=="sort") %>% 
  ggplot(., aes(x = cells_loaded, y = recovery, fill = recovery_type)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  theme_classic() +
  labs(
    x = "Cells Loaded",
    y = "Cell Recovery",
    fill = "Recovery Type") +
  scale_fill_manual(values = c("#8FAF8F","#5E548E"),
  labels = c("Targeted Recovery", "Actual Recovery")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title=element_text(size=20),
        axis.text=element_text(size=18),
        title=element_text(size=20),
        legend.text = element_text(size=18),
        legend.position=c(0.3,0.8)
        )

```

## Refine dataset for plotting

Note that we will remove a few clusters for being doublets, and we will refine cell type annotations across the dataset.

```{r umap}

ser$sample_name <- factor(
  ser$sample_name,
  levels = c("2.5K_Sort_2","5K_Sort_2", "10K_Sort_2", "20K_Sort_2"))

## extract UMAP and metadata for pretty plotting
umap <- Embeddings(ser,reduction="umap_int")
meta <- ser@meta.data

umap_meta <- cbind(umap,meta)
colnames(umap_meta)[1:2] <- paste("UMAP_",1:2,sep="")
umap_meta <- umap_meta %>%
  filter(!grepl("9|11",RNA_snn_res.0.3)) %>%
  mutate(refined_sample_names=recode(
    sample_name,
    `2.5K_Sort_2`="2,500 loaded cells",
    `5K_Sort_2`="5,000 loaded cells",
    `10K_Sort_2`="10,000 loaded cells",
    `20K_Sort_2`="20,000 loaded cells"
  )) %>%
  mutate(`Cells loaded`=recode(
    sample_name,
    `2.5K_Sort_2`="2,500",
    `5K_Sort_2`="5,000",
    `10K_Sort_2`="10,000",
    `20K_Sort_2`="20,000"
  )) %>%
  mutate(`Cell types`=cluster_anno) %>%
  mutate(`Cell types`=recode(`Cell types`,
    `CD8+ T Cells`="T cells",
    `cDC2`="Dendritic cells",
    `Mature B Cells`="B cells",
    `Naive B Cells`="B cells",
    `NK Cells`="NK cells",
    `Plasma Cells`="B cells"
  )) %>%
  mutate(`Cell types`=ifelse(RNA_snn_res.0.3==8 & UMAP_1< -5, "pDCs",`Cell types`)) %>% 
  mutate(Clusters=RNA_snn_res.0.3) %>%
  mutate(Clusters=droplevels(Clusters)) %>%
  mutate(Clusters=as.factor(as.numeric(Clusters)))

```

## UMAPs from chip sorting - cell titration and cluster frequencies

```{r}

p2 <- umap_meta %>%
  ggplot(.) +
  geom_point(aes(x=UMAP_1,y=UMAP_2),colour="black",size=1.2) +
  geom_point(aes(x=UMAP_1,y=UMAP_2,colour=Clusters),size=0.8) +
  scale_colour_manual(values=colors_pal) +
  theme_bw() +
  theme(
    panel.background = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    strip.background = element_blank(),
    strip.text = element_text(size=18),
    axis.title = element_text(size=20),
    legend.text = element_text(size=18),
    legend.title = element_text(size=20),
    legend.position = "bottom",
    legend.box.background = element_rect(color = "black", linewidth = 1)
  ) +
  guides(colour = guide_legend(override.aes = list(size=1.2))) +
  facet_wrap(~refined_sample_names)

p3 <- umap_meta %>%
  select(`Cells loaded`,Clusters) %>%
  group_by(`Cells loaded`) %>%
  mutate(total_cells=n()) %>%
  group_by(`Cells loaded`,Clusters) %>%
  mutate(cells_per_clust=n()) %>%
  ungroup() %>%
  mutate(`Cell proportion`=cells_per_clust/total_cells) %>%
  ggplot(.,aes(x=`Cells loaded`,y=`Cell proportion`,fill=Clusters)) +
  geom_col(position="fill") +
  scale_fill_manual(values=colors_pal) +
  xlab("Cells loaded") +
  ylab("Proporition of cluster in a sample") +
  scale_y_continuous(expand = expansion(mult = c(0, 0))) +
  theme_bw() +
  theme(
    panel.background = element_blank(),
    panel.grid = element_blank(),
    axis.title = element_text(size=20),
    axis.text = element_text(size=18),
    legend.text = element_text(size=18),
    legend.title = element_text(size=20),
    legend.position = "bottom",
    legend.box.background = element_rect(color = "black", linewidth = 1),
    axis.text.x=element_text(angle=45,hjust=T)
  )

```

## UMAPs from chip sorting - cell titration and cell type frequencies

```{r stacked bar plot}

p4 <- umap_meta %>%
  ggplot(.) +
  geom_point(aes(x=UMAP_1,y=UMAP_2),colour="black",size=1.2) +
  geom_point(aes(x=UMAP_1,y=UMAP_2,colour=`Cell types`),size=0.8) +
  scale_colour_manual(values=my_colors) +
  theme_bw() +
  theme(
    panel.background = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    strip.background = element_blank(),
    strip.text = element_text(size=18),
    axis.title = element_text(size=20),
    legend.text = element_text(size=10),
    legend.title = element_text(size=12),
    legend.position="bottom",
    legend.box.background = element_rect(color = "black", linewidth = 1)
  ) +
  guides(colour = guide_legend(nrow=3,override.aes = list(size=1.2))) +
  facet_wrap(~refined_sample_names)

p5 <- umap_meta %>%
  select(`Cells loaded`,`Cell types`) %>%
  group_by(`Cells loaded`) %>%
  mutate(total_cells=n()) %>%
  group_by(`Cells loaded`,`Cell types`) %>%
  mutate(cells_per_clust=n()) %>%
  ungroup() %>%
  mutate(`Cell proportion`=cells_per_clust/total_cells) %>%
  ggplot(.,aes(x=`Cells loaded`,y=`Cell proportion`,fill=`Cell types`)) +
  geom_col(position="fill") +
  scale_fill_manual(values=my_colors) +
  ylab("Proporition of cluster in a sample") +
  scale_y_continuous(expand = expansion(mult = c(0, 0))) +
  theme_bw() +
  theme(
    axis.title.x=element_blank(),
    panel.background = element_blank(),
    panel.grid = element_blank(),
    axis.title = element_text(size=20),
    axis.text = element_text(size=18),
    legend.text = element_text(size=10),
    legend.title = element_text(size=12),
    legend.position = "bottom",
    legend.box.background = element_rect(color = "black", linewidth = 1),
    axis.text.x=element_text(angle=45,hjust=TRUE)
  )

```

## Create plot

```{r}

p1
p2
p3
p4 
p5

```

## Session info

```{r}

sessionInfo()

```

